# Regression and model validation


```{r}
date()


```

## Load data

```{r}

library(dplyr)
library(tidyverse)
library(GGally)
library(ggplot2)
learning2014 <- read_csv('data/learning2014.csv')
str(learning2014)
```


This is a dataset collected from an exam. It has 166 observations (row) and 7 variables (col). The variables are `gender`, `age`, `attitude`, `deep`, `surf`, `stra`, and `points`. The value of these variables are mostly numbers, either integer or float, except gender, which is filled with two types of characters: "F" and "M". 

## Graphical overview and variable summary

```{r}

p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

p

```

With the code above, we are able to display the correlation between variables. As we can see from the figure, `attitude` and `points` have the most significant correlation. I suppose this makes sense in general.  Both `stra` and `surf` also have a strong correlation but `surf` has a negative correlation.

## Fit a linear model and its summary

```{r}

model <- lm(points ~ attitude + stra + surf, data = learning2014)

summary(model)


```

<!-- Show a summary of the fitted model and comment and interpret the results. Explain and interpret the statistical test related to the model parameters-->
As what is shown in the previous figure, I select `attitude`, `stra` and `surf` to be the explanatory variable. According to the summary, the coefficient for `attitude`, `stra` and `surf` is  3.3952, 0.8531 and -0.5861 respectively. The intercept of this model is 11.0171. 

The test of this model shows that the residual is distributed within -17.1550 and 10.9852 and the mean of residual is 0.5156. 


## Explain and interpret the multiple R-squared of the model
<!-- Using a summary of your fitted model, explain the relationship between the chosen explanatory variables and the target variable (interpret the model parameters). Explain and interpret the multiple R-squared of the model. -->
The summary of this model shows that `attitude` is the most significant variable in this model (Pr=1.93e-08), while the other two are less significant (Pr>0.01). This aligns with the coefficients of these three explanatory variables, where the coefficient of  `attitude` is the highest absolute value.

The multiple R-squared of this model is 0.207. It is not a high value, which suggests the model is not fully explainable by `attitude`, `stra`, and `surf`. Considering I am using explanatory variables that are most correlated with `points`, it means we either need to consider a more complicated model rather than simple linear regression or there are other missing explanatory variables that are not available from the dataset.


## Diagnostic plots
<!-- Produce the following diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage. Explain the assumptions of the model and interpret the validity of those assumptions based on the diagnostic plots. -->

The assumption of the simple linear model is that all explanatory variables follow a linear relationship with the target variable. The observations and residual should be independent. The residual should also follow the normal distribution. 


* Residuals vs Fitted: From this plot, we can see the red curve fluctuate around the perfect horizontal line. Scatter dots #35, #56 and #145 seem to be outliers in this plot.
```{r}
plot(model, which = 1)
```


* Normal QQ-plot: In this plot, we can see the majority of the dots fall along the straight diagonal line. There are a small number of observations that deviate from the line. #35, #56, and #145 seem to deviate the most. We can claim that the residual mostly follows the normal distribution.
```{r}
plot(model, which = 2)
```


* Residuals vs Leverage: This plot is supposed to show influential observations according to Cook's distance. However, I don't there is a curve for this metric. I suppose it means there aren't any influential observations in this dataset.
```{r}
plot(model, which = 5)
```


In conclusion, even though the model is not fully explainable by `attitude`, `stra`, and `surf`. The plots here suggest it is still appropriate to assume these explanatory variables follow a linear patterns in terms of the target variable.