
# Analysis of longitudinal data



```{r}
# This is a so-called "R chunk" where you can write R code.

date()
set.seed(0)
library(MASS)
library(dplyr)
library(tidyverse)
library(corrplot)
library(ggplot2)
library(lme4)
library(patchwork)
```


## RATS
 <!-- Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations) -->

### Load Data

I load RATS data from my csv dump, which is created from `meet_and_repear.R`. A summary of the original dataset `RATS` and its "long" form `RATSL` have been presented below:

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep="", header=TRUE)
glimpse(RATS)
RATSL <- read.csv('data/ratsl.csv', colClasses = c("ID" = "character", "Group" = "character"))
glimpse(RATSL)
```



### Graphical overview

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

The plot above is a simple overview for all 16 rats. They are plotted according to their group. We can observe that the weight of all rats is increasing across the time. We can also see tracking phenomenon from this plot, where a rat with a greater weight also will ended with a greater weight. Overall, the weight of rats from different group are generally close, excpet there seems to be an outlier in Group 2.

### Standardized overview

```{r}

# Standardise the variable weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised weight
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")

```

The plot above is based on standardized data. As we can observe from this plot, lines are much flatter if compared with the previous plot, which we can see the tracking phenomenon more clearly.

### Average profile by groups

```{r}

# Summary data with mean and standard error of Weight by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")

```

The plot above shows the average profiles for each group. We can observe that the weight of rats in Group 1 is lowest on average, while rats in Group 3 have the highest weight on average. We can also see rats in Group 2 and Group 3 shows more variability throughout the time than rats in Group 1.


### Mean summary measure

```{r}

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 1)
RATSLS <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSLS)

# Draw a boxplot of the mean versus Group
ggplot(RATSLS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")


```

The plot above is a boxplot that shows the mean summary measure. It shows that the mean summary measure is more variable in Group 2. According to this plot, all of the groups seem to have an outlier. After filtering all outliers, a new plot is presented below:

```{r}

RATSLS1 <- filter(RATSLS, 
                    mean > 250 & Group == '1' | 
                    mean < 500 & Group == '2' | 
                    mean > 500 & Group == '3')
ggplot(RATSLS1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")
```

### T-test and Anova

Despite all the plots above shows that these three group are quite different, I conducted t-test for these three group of data in pair anyway. The result is printed below:

```{r}
for(i in list(c('1', '2'), c('2', '3'), c('1', '3'))){
    print(t.test(mean ~ Group, 
    data = filter(RATSLS1, Group == i[1] | Group == i[2]), 
    var.equal = TRUE))
}
```

The result here formally confirms that all the groups are different from each other with all the p-value lower than 0.05. 

I also conducted anova test below:

```{r}

# Add the baseline from the original data as a new variable to the summary data
RATSLS2 <- RATSLS %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSLS2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

As we can observe from this result, the baseline (`WD1`) has the most significant correlation with weight during the experiment. We can also observe that `Group` also has a strong correlation with the weight. It suggests the evidence of a diat difference among different group.

## BPRS
 <!-- Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations) -->

### Load data

I load BPRS data from my csv dump, which is created from `meet_and_repear.R`. A summary of the original dataset `BPRS` and its "long" form `BPRSL` have been presented below:

```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)
glimpse(BPRS)
BPRSL <- read.csv('data/bprsl.csv', colClasses = c("treatment" = "character", "subject" = "character"))
glimpse(BPRSL)
```

### Graphical overview

```{r}

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

The plot above is a visualization of the change of weights of all 40 people in the dataset `BPRSL`. They are plotted according to the treatment group. 


The plot below is a scatterplot matrix. As the MABS book suggests, it is not really helpful, but it can show that repeated measures are dependent on one another.

```{r}
pairs(dplyr::select(RATS, -ID, -Group))
```


### Linear mixed effect models

#### Random intercept model

I first fit a regular linear regression model. This model assumes independence. Its summary is printed below:

```{r}
# create a regression model
BPRS_reg <- lm(bprs ~ week + treatment, BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

As what the summary suggests, `week` are the most significant variable in this model. Next, I fit a random intercept model models, which allows within-subject dependencies. Again, its summary is printed below:

```{r}

BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

Comparing these two summary, we can see the estimate and standard error are all very close, which is in conflict with the MABS book: assuming independence will lead to `treatment` being more precise and `week` being larger. But the summary also shows that the `treatment` is not significant at all. This may suggest that there is no real relationship between the target variable and the predictor variable `treatment`. 

#### Random intercept and random slope model

I fit *Random Intercept and Random Slope Model* according to MABS book and conducted an anova test. The result is printed below:

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
# print a summary of the model
summary(BPRS_ref1)
# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

The test result shows that random intercept and slope model gives a chi-squared statistic of 7.27 with 2 degrees of freedom. Its associated p-value is 0.02, which suggests random intercept and slope model can offer a better fit than the random intercept model.

#### Random intercept and random slope model with interaction

In the end, I fitted a random intercept and slope model that allows for a week x treatment interaction. The result is printed below:

```{r}

BPRS_ref2 <- lmer(bprs ~ week + treatment + week * treatment + (week | subject), data = BPRSL, REML = FALSE)
# print a summary of the model
summary(BPRS_ref2)
# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

```

The result of anova test above shows that the model with interaction can give a chi-squared statistic of 3.17 with 1 degree of freedom. Its associated p-value is not so small (0.07). This suggests that the mode with interaction may not fit better than the previous model which doesn't have interaction. 

```{r}
BPRSL <- BPRSL %>% mutate(Fitted = fitted(BPRS_ref2))

plot_a <- ggplot(BPRSL, aes(x = week, y = bprs, color = treatment)) +
  geom_line(aes(linetype = subject)) +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))


plot_b <- ggplot(BPRSL, aes(x = week, y = Fitted, color = treatment)) +
  geom_line(aes(linetype = subject)) +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))


(plot_a + plot_b  & theme(legend.position = "bottom")) + plot_layout(guides = "collect")
```


The plot above shows how well the random intercept and slope model with interaction fits. It is displayed side by side, where the plot of data is on the left side and the fitted model is placed on the right side. The `treatment` is visualized by color and the `subject` is visualized by linestyle. 

We can see there are too many lines cluttered in a very close space for both left and right sides. I suppose we cant tell too much from these two plots. But one obvious thing is that all subjects have their weight decreasing regardless of treatment. 